Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-02-09 paper.tex
before meeting

We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.  This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations. First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.  For example, the current model (see Methods) takes 19.3 seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes 24619 seconds to solve when $\alpha=10$. Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort. The second situation under which sensitivity to system scaling could become important is when measurements are relative and do not provide single-molecule resolution. For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).

It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{XXX van kampen? Gillespei look for this}. As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}. For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved. Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.

@@

When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. 
% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.  We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.

>>>>

We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.  This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations. First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.  For example, the current model (see Methods) takes 19.3 seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes 24619 seconds to solve when $\alpha=10$. Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort. The second situation under which sensitivity to system scaling could become important is when measurements are relative and do not provide single-molecule resolution. For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}). 
% MMM - Ding:2019??
It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{Kampen1961}. As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}. For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved. Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.

@@

When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $k_0$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. 
% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
% MMM - This was k_0, not eta
In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.  We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.
